---
title: "NYC_Italian_restaurant"
author: "Saahil Shroff"
date: "2022-12-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(tidyverse)
library(dplyr)
```


# Find Cheap Restaurants
## Loading the data and performing basic sanity checks.
```{r}
load_nyc_restaurants_df = read.delim("./nyc-italian-cheap.csv.bz2")
head(load_nyc_restaurants_df)

cat("Dimensions on the loaded dataframe:",dim(load_nyc_restaurants_df))
cat("Are there any NULLs present in the loaded data frame?", any(is.na(load_nyc_restaurants_df)))
```

```{r}
cat("Logistic regression is suitable for predicting if a restaurant is cheap or not. It is because logistic regression is used for classification tasks since it is a simple and effective way to predict a binary outcome. The model uses logistic function to model the relationship between a binary dependent variable (cheap variable in this case) and one or more independent variables (such as decor, service) to predict the probability that an instance belongs to a certain class. Linear regression, on the other hand, is used for predicting continuous values, and would have been suitable for predictiong price of a meal at a restaurant.")
```

## Building the model - estimating the model and interpreting the results.

#### Factoring food,decor,and service using the following categories:
#### 0-4: 0; 
#### 5-9: 1;  
#### 10-14: 2; 
#### 15-19: 3; 
#### 20-24: 4; 
#### 25 =<: 5.
```{r}
split_food = cut(load_nyc_restaurants_df$Food, 
                breaks = c(0,5,10,15,20,25,Inf),
                labels=c("0","1", "2", "3", "4", "5"))
split_decor = cut(load_nyc_restaurants_df$Decor, 
                breaks = c(0,5,10,15,20,25,Inf),
                labels=c("0","1", "2", "3", "4", "5"))
split_service = cut(load_nyc_restaurants_df$Service, 
                breaks = c(0,5,10,15,20,25,Inf),
                labels=c("0","1", "2", "3", "4", "5"))

nyc_restaurants = data.frame(load_nyc_restaurants_df,split_food,split_decor,split_service)
head(nyc_restaurants)
```
#### Creating a logistic model:
```{r}
restaurant_model = glm(Cheap ~ factor(split_food) + factor(split_decor) + factor(split_service) + factor(East), data=nyc_restaurants, family="binomial")
summary(restaurant_model)
```
```{r}
cat("My assumption was that all the variables - Food, Decor, Service, and East will play a role to determine if a restaurant is cheap or not. In my experience, if the food, decor, service are top-class and it is situated in a posh locality, there are high chances for the restaurant to be expensive. However, looking at the model, only top-class food (4/5) is statistically significant towards prediciting if a restaurant is cheap or not, which goes against my common observation and experience.")
```

## Adding two new places with the following scores and locations to make predictions
```{r}
# Creating dataframe on which prediction has to be made 
Restaurant = c("Assagio Ristorante","Altura")
Food = c(23,18)
Decor = c(17,15)
Service = c(22,24)
East = c(0,1)
to_predict = data.frame(Restaurant,Food,Decor,Service,East)
to_predict
```
#### Getting the dataframe ready for prediction:
```{r}
split_food = cut(to_predict$Food, 
                breaks = c(0,5,10,15,20,25,Inf),
                labels=c("0","1", "2", "3", "4", "5"))
split_decor = cut(to_predict$Decor, 
                breaks = c(0,5,10,15,20,25,Inf),
                labels=c("0","1", "2", "3", "4", "5"))
split_service = cut(to_predict$Service, 
                breaks = c(0,5,10,15,20,25,Inf),
                labels=c("0","1", "2", "3", "4", "5"))
# Adding the classification to the "prediction" data frame
to_predict = data.frame(to_predict,split_food,split_decor,split_service)
to_predict

library(caret)
cheap_restaurant_prediction = predict(restaurant_model,to_predict)
# cheap_restaurant_prediction
normalized_prediction = ifelse(cheap_restaurant_prediction>=0.5,1,0) # grouping together for classification
final_prediction = cbind(to_predict,normalized_prediction)
```
#### Prediction and Inference:
```{r}
final_prediction %>% select(Restaurant,normalized_prediction)
cat("Thus, according to the logistic-regression model's predicition, Altura is a cheap Italian restaurant, while Assagio Ristorante is an expensive Italian restaurant.")
```
